---
title: "species_for_mocks"
author: "Eily Allan"
date: "11/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

Let's take the combined ASV tables for runs 1-4 and find the top (and bottom?) hashes for each marker. Maybe take the top 10 unique species and top 10 bottom species to put in - that way we can tell if species are not amplifying because they have a low amplification efficiency (a) or if they just really aren't in our samples. 

After that, we will add in all the salmonids and a few other target species of interest (for fish - bull trout, lamprey, etc.)

```{r import files}

# add in the combined ASV tables from runs 1-4
all.coi.asvs <- read.csv("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/dada2_output/20211120.combined.COI.ASV.table.csv")
all.mifish.asvs <- read.csv("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/dada2_output/20211120.combined.MiFish.ASV.table.csv")
all.mimammal.asvs <- read.csv("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/dada2_output/20211120.combined.MiMammal.ASV.table.csv")

# now we also need the classifications to see what the hashes actually are 
# i think we can use the most recent genus and species classifications (because we need to know those to put in)
all.coi.classifications <- readRDS("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/classification_output/COI.all.gs.previous.hashes.annotated.rds")
all.coi.classifications <- all.coi.classifications %>% 
  rename(Hash = representative)

all.mifish.classifications <- readRDS("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/classification_output/MiFish.all.gs.previous.hashes.annotated.rds")
all.mifish.classifications <- all.mifish.classifications %>% 
  rename(Hash = representative)

all.mimammal.classifications <- readRDS("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/classification_output/MiMammal.all.gs.previous.hashes.annotated.rds")
all.mimammal.classifications <- all.mimammal.classifications %>% 
  rename(Hash = representative)

# lets add in the blast results too in case there are some that weren't in here
all.coi.classifications2 <- readRDS("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/classification_output/COI.blast.previous.hashes.annotated.rds")
all.coi.classifications2 <- all.coi.classifications2 %>% 
  rename(Hash = representative)

all.mifish.classifications2 <- readRDS("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/classification_output/MiFish.blast.previous.hashes.annotated.rds")
all.mifish.classifications2 <- all.mifish.classifications2 %>% 
  rename(Hash = representative)

all.mimammal.classifications2 <- readRDS("/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/classification_output/MiMammal.blast.previous.hashes.annotated.rds")
all.mimammal.classifications2 <- all.mimammal.classifications2 %>% 
  rename(Hash = representative)

# then check to see if there are any new hashes in blast that aren't in the previous 
intersect(all.coi.classifications$Hash, all.coi.classifications2$Hash)
intersect(all.mifish.classifications$Hash, all.mifish.classifications2$Hash)
intersect(all.mimammal.classifications$Hash, all.mimammal.classifications2$Hash)

#none so phew
```

```{r attach taxonomy}

coi.all <- all.coi.asvs %>% 
  group_by(Hash) %>% # for each sample that has multiple asvs that assign to the same taxa...
  summarise(tot = sum(nReads)) %>% 
  arrange(desc(tot)) %>% 
  left_join(all.coi.classifications, by= "Hash") 

# for coi select only metazoans
coi.metazoan <- coi.all %>% 
  filter(kingdom == "Metazoa")

coi.species <- coi.metazoan %>% 
  filter(species != "") %>% 
  arrange(desc(tot))

mifish.all <- all.mifish.asvs %>% 
  group_by(Hash) %>% # for each sample that has multiple asvs that assign to the same taxa...
  summarise(tot = sum(nReads)) %>% 
  arrange(desc(tot)) %>% 
  left_join(all.mifish.classifications, by= "Hash")

mifish.species <- mifish.all %>% 
  filter(species != "") %>% 
  arrange(desc(tot))

mimammal.all <- all.mimammal.asvs %>% 
  group_by(Hash) %>% # for each sample that has multiple asvs that assign to the same taxa...
  summarise(tot = sum(nReads)) %>% 
  arrange(desc(tot)) %>% 
  left_join(all.mimammal.classifications, by= "Hash")

mimammal.species <- mimammal.all %>% 
  filter(species != "") %>% 
  arrange(desc(tot))

```

```{r}

# there arent that many for 12s ... lets just get all unique species

mifish.unique <- unique(mifish.species$species)
mimammal.unique <- unique(mimammal.species$species)
coi.unique <- unique(coi.species$species)
#write.csv(coi.unique, "/Users/elizabethandruszkiewicz/Desktop/coi.metazoans.to.species.csv")


# ok lets take marker, nReads, taxon to make a table for ryan
coi.simple <- coi.species %>% 
  select(tot, taxon)
coi.simple$marker = "COI"
  
mifish.simple <- mifish.species %>% 
  select(tot, taxon)
mifish.simple$marker = "MiFish"

mimammal.simple <- mimammal.species %>% 
  select(tot, taxon)
mimammal.simple$marker = "MiMammal"

all.detected.species.level <- rbind(coi.simple, mifish.simple, mimammal.simple)
write.csv(all.detected.species.level, "/Users/elizabethandruszkiewicz/GoogleDrive/UW/GitHub/NextGenNEPA_EA/Output/all.detected.species.level.csv")

```

```{r compare markers}
unique.markers <- 
  all.detected.species.level %>% 
  group_by(taxon, marker) %>% 
  summarize(total= sum(tot))

unique.num.markers <- 
  unique.markers %>% 
  count(taxon)

mix.plot <- 
  unique.markers %>% 
  filter(marker != "COI")

# bar plot - this is bad and also too many taxa
ggplot(test, aes(fill=marker, y=total, x=taxon)) + 
    geom_bar(position="dodge", stat="identity")

# heat map of only mifish and mimammal 
ggplot(mix.plot, aes(marker, taxon, fill= log10(total))) + 
  geom_tile() +
  theme_bw() +
  scale_fill_gradient(low="white", high="blue")

```

```{r more plotting}

# now lets sum by family to compare three markers
coi.family <- coi.species %>%
  group_by(class) %>% 
  summarize(total = sum(tot))
coi.family <- coi.family %>% 
  filter(class !="")
coi.family$marker = "COI"

mifish.family <- mifish.species %>%
  group_by(class) %>% 
  summarize(total = sum(tot))
mifish.family <- mifish.family %>% 
  filter(class !="")
mifish.family$marker = "MiFish"

mimammal.family <- mimammal.species %>%
  group_by(class) %>% 
  summarize(total = sum(tot))
mimammal.family <- mimammal.family %>% 
  filter(class !="")
mimammal.family$marker = "MiMammal"

all.detected.family.level <- rbind(coi.family, mifish.family, mimammal.family)

# let's try heat map now of all three markers but at family level
ggplot(all.detected.family.level, aes(marker, class, fill= log10(total))) + 
  geom_tile() +
  theme_bw() +
  scale_fill_gradient(low="white", high="blue")

#### STILL TOO CRAZY - go back to order - right now too lazy so just change it but not variable names

```

```{r with creek level stuff}
mix.together <- rbind(mifish.species, mimammal.species)


```