---
title: "add_taxonomy_to_samples"
author: "Eily Allan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

Last updated: 2/10/22

# Overview

After merging all the ASV tables and hash keys, let's pair them together with taxonomy - and also go from ASV tables to taxa tables and collapse to different taxonomic ranks depending on what types of analyses we want to do. 

# Set up

## Load libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(vegan)
library(reshape2)
```

## Read in files for all ASV tables, annotations, and metadata. 

We probably just want "good" annotations, which is family level annotations for COI and species level annotations for 12S. If you want to look at classification to any level, change the file paths for the classification outputs to all hashes. 

```{r read in files}

# Read the merged ASV tables
COI.ASV.table <- read_csv(paste0(here("Output", "dada2_output", "20220210.combined.COI.ASV.table.csv")))
MiFish.ASV.table <- read_csv(paste0(here("Output", "dada2_output", "20220210.combined.MiFish.ASV.table.csv")))
MiMammal.ASV.table <- read_csv(paste0(here("Output", "dada2_output", "20220210.combined.MiMammal.ASV.table.csv")))

# Read in the classification output for each run 
# the files written to the general folder (not run specific) should have everything classified from all runs because they are our "databases" to use for future runs - we want to use the "DATE.MARKER.all.good.previous.hashes.annotated.rds" file 
COI.annotations <- readRDS(file=paste0(here("Output","classification_output"),"/COI.all.gs.previous.hashes.annotated.rds"))
MiFish.annotations <- readRDS(file=paste0(here("Output","classification_output"),"/MiFish.all.gs.previous.hashes.annotated.rds"))
MiMammal.annotations <- readRDS(file=paste0(here("Output","classification_output"),"/MiMammal.all.gs.previous.hashes.annotated.rds"))

all.metadata <- read.csv(here("Input", "sequencing_metadata_files", "master_sequencing_datasheet_20211026.csv"))

```

# Combine ASV tables with taxonomy files 

We need to do this for each marker. 

## MiMammal

```{r ASV and taxonomy for MiMammal}

# this taxonomy file is the "good" annotations to species and genus level so let's get rid of the other junk for right now 
MiMammal.simple.annotations <- MiMammal.annotations %>% select(representative, taxon, species) %>% rename(Hash = representative)

MiMammal.ASV.taxon <- MiMammal.ASV.table %>% 
  left_join(MiMammal.simple.annotations, by = "Hash") %>% 
  select(-Locus)

# and same - keep things that do have an annotation (reminder to species or genus level so "good" annotation)
MiMammal.ASV.yes.taxon <- MiMammal.ASV.taxon[! is.na(MiMammal.ASV.taxon$taxon),]

# we can also do this at the species level 
MiMammal.by.species <- MiMammal.ASV.yes.taxon %>% 
  select(-Hash) %>% # we don't need the hash identifier anymore
  filter(species != "") %>% 
  group_by(Sample_name, species) %>% # for each sample that has multiple asvs that assign to the same taxa...
  summarise(tot = sum(nReads))

write.csv(MiMammal.by.species, here("Output","combined_species_tables","20220210_MiMammal.csv"))
```

```{r MiFish}

# this taxonomy file is the "good" annotations to species and genus level so let's get rid of the other junk for right now 
MiFish.simple.annotations <- MiFish.annotations %>% select(representative, taxon, species) %>% rename(Hash = representative)

MiFish.ASV.taxon <- MiFish.ASV.table %>% 
  left_join(MiFish.simple.annotations, by = "Hash") %>% 
  select(-Locus)

# and same - keep things that do have an annotation (reminder to species or genus level so "good" annotation)
MiFish.ASV.yes.taxon <- MiFish.ASV.taxon[! is.na(MiFish.ASV.taxon$taxon),]

# we can also do this at the species level 
MiFish.by.species <- MiFish.ASV.yes.taxon %>% 
  select(-Hash) %>% # we don't need the hash identifier anymore
  filter(species != "") %>% 
  group_by(Sample_name, species) %>% # for each sample that has multiple asvs that assign to the same taxa...
  summarise(tot = sum(nReads))

write.csv(MiFish.by.species, here("Output","combined_species_tables","20220210_MiFish.csv"))

```

```{r COI}

# this taxonomy file is the "good" annotations to species and genus level so let's get rid of the other junk for right now 
COI.simple.annotations <- COI.annotations %>% select(representative, taxon, species) %>% rename(Hash = representative)

COI.ASV.taxon <- COI.ASV.table %>% 
  left_join(COI.simple.annotations, by = "Hash") %>% 
  select(-Locus)

# and same - keep things that do have an annotation (reminder to species or genus level so "good" annotation)
COI.ASV.yes.taxon <- COI.ASV.taxon[! is.na(COI.ASV.taxon$taxon),]

# we can also do this at the species level 
COI.by.species <- COI.ASV.yes.taxon %>% 
  select(-Hash) %>% # we don't need the hash identifier anymore
  filter(species != "") %>% 
  group_by(Sample_name, species) %>% # for each sample that has multiple asvs that assign to the same taxa...
  summarise(tot = sum(nReads))

write.csv(COI.by.species, here("Output","combined_species_tables","20220210_COI.csv"))

```
