---
title: "explore_delta_euks"
author: "Eily Allan"
date: "2/10/2022"
output: html_document
---

Last modified 2/10/22


# Overview

Paper 2 explores the changes in eukaryotes *due to* culvert replacement. Here we are going to start exploring ways to do that. Basically here we want to see changes in community (as measured by our COI marker) in the target creek - Padden - and see how the culver replacement impacted these changes. 

# Set Up

## Load packages

```{r set up}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(vegan)
library(reshape2)

```

## Import data

For now, let's just focus on presence/absence data from our COI marker. Later we can go back and do COI/COI (or try to make it quantitative if we can get the variable PCR model working.)

We also have some finagling to do like removing technical replicates and positive controls (and Maya's samples). 

```{r import data}

COIdata <- read.csv(here("Output","combined_species_tables","20220210_COI.csv"))

# what if we use ASVs instead of species
COIdata <- read.csv(here("Output","dada2_output","20220210.combined.COI.ASV.table.csv"))

COI <- COIdata %>% 
  #select(-X) %>% 
  #select(-Locus) %>% 
  separate(Sample_name, into=c("Marker","Date","Creek","Site","Bio_Rep","extra"), extra="warn", remove=FALSE) %>% 
  filter(is.na(extra)) %>% select(-extra) %>% # remove the "Barnes Up2 old" sample - shouldn't normally need this
  filter(Bio_Rep == "1" | Bio_Rep == "2" | Bio_Rep == "3") %>%  # things that don't have bio reps are kangaroo or maya's samples
  filter(Creek != "Kangaroo")  # still some leftover kangaroos.. whoops 

```

# Data exploration! 

Ok! What do we find?? Let's start by plotting some stuff and then figure out what we are working with. 

## But first - kangaroo 

How well did we do avoiding cross-contamination and tag jumping?

```{r check kangaroo}
COI_kangaroo <- COIdata %>% 
  select(-X) %>% 
  filter(str_detect(Sample_name, "Kangaroo"))

# Wow. Only Kangaroo in there. Pretty amazing.. now quick check to see if the environmental samples have any kangaroo in them

COI_kangaroo_in_samples <- COI %>% 
  filter(str_detect(species, "Macropus") | str_detect(species, "Osphranter"))

# Nothing! Yahoo! 
```

## Richness of each creek over time

```{r creeks over time}

# this gives us the number of biological replicates that have a detection (presence/absence) for each species - at each site (up/down) within each creek (all 5) at each time point 

bysite <- COI %>% 
  group_by(Date, Creek, Site) %>%
  tally(Hash)
  #count(species)

asv_in_1 <- bysite %>% 
  filter(n==1)
asv_in_2 <- bysite %>% 
  filter(n==2)
asv_in_3 <- bysite %>% 
  filter(n==3)

# check to make sure that none of the samples have more than 3 (biological replicates)
sum(bysite$n >3)

# check richness at each time/site/creek 
rich <- bysite
rich$n <- 1

richness <- rich %>% 
  group_by(Date, Creek, Site) %>%
  summarize(richness = sum(n))

# plot it
ggplot(richness, aes(x=Date, y=richness, color=Site)) +
  facet_wrap(~Creek) +
  geom_point() +
  theme_bw() + 
  labs(title="COI: ASVs found in any replicate", y="Number of unique ASVs", x= "Month")


richness_all3 <- bysite %>% 
  filter(n == 3) %>% 
  group_by(Date, Creek, Site) %>%
  summarize(richness = sum(n)/3)

# plot it
ggplot(richness_all3, aes(x=Date, y=richness, color=Site)) +
  facet_wrap(~Creek) +
  geom_point() +
  theme_bw() +
  labs(title="Found in all 3 replicates", y="Number of unique ASVs", x= "Month")

```

## Difference in richness upstream and downstream 

```{r delta richness}
delta_rich <- richness %>% 
  group_by(Date,Creek) %>% 
  pivot_wider(names_from = Site, values_from = richness)

richplot <- delta_rich %>% 
  filter(Creek != "4Pad") %>% 
  mutate(rich_up_minus_down = Up - Dn) %>% 
  select(c(Date,Creek,rich_up_minus_down))

# plot it
controlsbox <- ggplot(richplot, aes(x=Date, y=rich_up_minus_down)) +
  geom_boxplot(fill="#00BFC4") +
  theme_bw() +
  labs(title="COI: All Control Creeks", y="ASV Richness Upstream - ASV Richness Downstream", x= "Month")

# now go back and do padden
padden_richplot <- delta_rich %>% 
  filter(Creek == "4Pad") %>% 
  mutate(rich_up_minus_down = Up5 - Dn) %>% 
  select(c(Date,Creek,rich_up_minus_down))

# plot it
interventionbox <- ggplot(padden_richplot, aes(x=Date, y=rich_up_minus_down)) +
  geom_boxplot(fill="#F8766D") +
  theme_bw() +
  labs(title="COI: Intervention Creek", y="ASV Richness Upstream - ASV Richness Downstream", x= "Month")

controlsbox + interventionbox + plot_layout(ncol = 2, nrow = 1, byrow = F)
```

## Different way of looking at richness

What if instead of counting each unique thing in all three replicates towards one estimate of richness, instead we calculate three richnesses (one from each bio rep) and box plot it 

```{r different way of looking at richness}

new_rich <- COI %>% 
  group_by(Date, Creek, Site, Bio_Rep) %>%
  tally()

delta_new_rich <- new_rich %>% 
  group_by(Date,Creek,,Bio_Rep) %>% 
  pivot_wider(names_from = Site, values_from = n)

newrichplot <- delta_new_rich %>% 
  filter(Creek != "4Pad") %>% 
  mutate(rich_up_minus_down = Up - Dn) %>% 
  select(c(Date,Creek,rich_up_minus_down))

# plot it
newcontrolsbox <- ggplot(newrichplot, aes(x=Date, y=rich_up_minus_down)) +
  geom_boxplot(fill="#00BFC4") +
  theme_bw() +
  labs(title="COI: All Control Creeks (sep bio reps)", y="ASV Richness Upstream - ASV Richness Downstream", x= "Month")

# now go back and do padden
newpadden_richplot <- delta_new_rich %>% 
  filter(Creek == "4Pad") %>% 
  mutate(rich_up_minus_down = Up5 - Dn) %>% 
  select(c(Date,Creek,rich_up_minus_down))

# plot it
newinterventionbox <- ggplot(newpadden_richplot, aes(x=Date, y=rich_up_minus_down)) +
  geom_boxplot(fill="#F8766D") +
  theme_bw() +
  labs(title="COI: Intervention Creek (sep bio reps)", y="ASV Richness Upstream - ASV Richness Downstream", x= "Month")

newcontrolsbox + newinterventionbox + plot_layout(ncol = 2, nrow = 1, byrow = F)

```

## NMDS plots for sanity check    

Biological replicates really don't match up well. So do things even cluster by creek, month, and station? 

```{r nmds}

coi.asv.for.nmds <- COI %>% 
  #select(-Locus) %>% 
  #select(c(Sample_name,species,tot)) %>% 
  select(c(Sample_name,species,nReads)) %>%
  pivot_wider(names_from = Sample_name, values_from = nReads) %>%
  #pivot_wider(names_from = Sample_name, values_from = tot) %>% 
  column_to_rownames(var="Hash")
  #column_to_rownames(var="species")

coi.asv.for.nmds[is.na(coi.asv.for.nmds)] <- 0
  
coi.asv.nmds <- metaMDS(t(coi.asv.for.nmds), distance = "bray")

goodness(coi.asv.nmds) # Produces a results of test statistics for goodness of fit for each point
stressplot(coi.asv.nmds) # Produces a Shepards diagram
plot(coi.asv.nmds, "sites") 
  
data.scores = as.data.frame(scores(coi.asv.nmds))
meta <- unique(COI[,1:6])

data.scores$Sample = meta$Sample_name
data.scores$Date = meta$Date
data.scores$Creek = meta$Creek


ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 4, aes(shape = Creek, fill= Date), colour = "black" )+ 
    labs(x = "NMDS1", fill = "Date", y = "NMDS2", shape = "Creek") +
    scale_shape_manual(values=21:25) +
  scale_fill_brewer(palette = "Pastel2") +
  guides(fill=guide_legend(override.aes=list(colour=c("0321"="azure","0421"="red", "0521"="black", "0621"="black", "0721"="black", "0821"="black"))))
 

```

