---
title: "explore_delta_euks"
author: "Eily Allan"
date: "2/10/2022"
output:
  pdf_document: default
  html_document: default
---

Last modified 2/10/22


# Overview

Paper 2 explores the changes in eukaryotes *due to* culvert replacement. Here we are going to start exploring ways to do that. Basically here we want to see changes in community (as measured by our COI marker) in the target creek - Padden - and see how the culver replacement impacted these changes. 

# Set Up

## Load packages

```{r set up}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(tidyverse)
library(vegan)
library(reshape2)
library(patchwork)
```

## Import data

Determine which marker and whether analyzing at species or ASV level.

```{r which data}

## must choose if you want to analyze at species or ASV level
datatype = "species" 
#datatype = "ASV"

## must choose marker 
marker = "COI"

```

Now read it in and format it 

```{r import data}

# change to most recent date when files were created
mostrecent <- "20220210"

##
if(datatype=="species"){
  filename <- paste0(mostrecent,"_",marker,".csv")
  filepath <- here("Output","combined_species_tables",filename)
  rawdata <- read.csv(filepath)
  data <- rawdata %>% 
    select(-X) %>% 
    rename(reads = tot) %>% 
    rename(species_asv = species) %>% 
    separate(Sample_name, into=c("Marker","Date","Creek","Site","Bio_Rep","extra"), extra="warn", remove=FALSE) %>% 
    filter(is.na(extra)) %>% select(-extra) %>% # remove the "Barnes Up2 old" sample - shouldn't normally need this
    filter(Bio_Rep == "1" | Bio_Rep == "2" | Bio_Rep == "3") %>%  # things that don't have bio reps are kangaroo or maya's samples
    filter(Creek != "Kangaroo")  # still some leftover kangaroos.. whoops 
}

if(datatype=="ASV"){
  filename <- paste0(mostrecent,".combined.",marker,".ASV.table.csv")
  filepath <- here("Output","dada2_output",filename)
  data <- read.csv(filepath)
  rawdata <- read.csv(filepath)
  data <- rawdata %>% 
    rename(reads = nReads) %>% 
    rename(species_asv = Hash) %>% 
    separate(Sample_name, into=c("Marker","Date","Creek","Site","Bio_Rep","extra"), extra="warn", remove=FALSE) %>% 
    filter(is.na(extra)) %>% select(-extra) %>% # remove the "Barnes Up2 old" sample - shouldn't normally need this
    filter(Bio_Rep == "1" | Bio_Rep == "2" | Bio_Rep == "3") %>%  # things that don't have bio reps are kangaroo or maya's samples
    filter(Creek != "Kangaroo")  # still some leftover kangaroos.. whoops 
}

```

We have 3 biological replicates for each site. Let's make three different versions that we can use to plot. One where we count only things found in all 3/3 replicates, one with anything found in at least 2 replicates, and one where we count all unique things found even if they were just in 1 replicate. This will give us a sense for how well our replicates match as well. 

```{r different numbers of replicates}
# this gives us the number of biological replicates that have a detection (presence/absence) for each species - at each site (up/down) within each creek (all 5) at each time point 
bysite <- data %>% 
  group_by(Date, Creek, Site) %>%
  count(species_asv)

# check to make sure that none of the samples have more than 3 (biological replicates)
sum(bysite$n >3)

# things found in any of replicates (1, 2, or 3/3)
data.123 <- data 

# things found in at least 2 replicates (2 or 3/3)
reps.23 <- bysite %>% 
  filter(n != 1) %>% 
  unite("id", c(Date,Creek,Site,species_asv), remove=FALSE)

data.23 <- data %>% 
  unite("id", c(Date,Creek,Site,species_asv), remove=FALSE) %>% 
  filter(id %in% reps.23$id)

# things found in all 3 replicates 
reps.3 <- bysite %>% 
  filter(n == 3) %>% 
  unite("id", c(Date,Creek,Site,species_asv), remove=FALSE)

data.3 <- data %>% 
  unite("id", c(Date,Creek,Site,species_asv), remove=FALSE) %>% 
  filter(id %in% reps.3$id)

```

# Metrics and plotting 

## Richness calculations

We want to do this for all the different versions (1,2,3 bio reps). And then we can compare in each sample how many things are found in 1, 2, or all 3 replicates. 

```{r richness calculations}

# check richness at each time/site/creek 

rich.123 <- data.123 %>% 
  select(c(Date,Creek,Site,species_asv)) %>% 
  group_by(Date, Creek, Site) %>%
  summarise(richness = n_distinct(species_asv))

rich.23 <- data.23 %>% 
  select(c(Date,Creek,Site,species_asv)) %>% 
  group_by(Date, Creek, Site) %>%
  summarise(richness = n_distinct(species_asv))

rich.3 <- data.3 %>% 
  select(c(Date,Creek,Site,species_asv)) %>% 
  group_by(Date, Creek, Site) %>%
  summarise(richness = n_distinct(species_asv))


biorep.richness <- rich.123 %>% 
  rename(richness123 = richness) %>% 
  left_join(rich.23, by = c("Date" = "Date", "Creek" = "Creek", "Site"="Site")) %>% 
  rename(richness23 = richness) %>% 
  left_join(rich.3, by = c("Date" = "Date", "Creek" = "Creek", "Site"="Site")) %>% 
  rename(richness3 = richness)

```


## Richness plots

```{r richness plots}

# plot difference in richness when including things found in 1/2/3 biological replicates

make.rich.biorep.plot <- function(df, numreps){
  ylabel <- paste0("Number of unique ", datatype)
  titlelabel <- paste0(marker, ": ", datatype, " found per site in at least ", as.character(numreps), "/3 biological replicates")
  p <- df %>% 
  ggplot(aes(x=Date, y= richness, fill=Site)) +
  facet_wrap(~Creek) +
  scale_fill_manual(values = c("#003333", "#CCFFFF", "#669999","#CCFFFF") ) +
  geom_bar(position="dodge", stat="identity", color="black") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45)) +
  labs(title= titlelabel, y= ylabel, x= "Month")
  
  return(p)
}

p1 <- make.rich.biorep.plot(rich.123, 1) + theme(legend.position = "none")
p1file = paste0(marker,"_",datatype,"_richness_atleast1biorep.png")
p2 <- make.rich.biorep.plot(rich.23, 2) + theme(legend.position = "none")
p2file = paste0(marker,"_",datatype,"_richness_atleast2biorep.png")
p3 <- make.rich.biorep.plot(rich.3, 3) 
p3file = paste0(marker,"_",datatype,"_richness_atleast3biorep.png")

#ggsave(p1, filename= here("Output","figures",p1file))
#ggsave(p2, filename= here("Output","figures",p2file))
#ggsave(p3, filename= here("Output","figures",p3file))

p1 + p2 + p3 + plot_layout(ncol = 3, nrow = 1, byrow = F)

```

## Difference in richness upstream and downstream 

Let's only do this for the big richness (speices found in any replicate)

```{r delta richness}

# here df is rich.123
calc.delta.rich <- function(df){
  
  p <- df %>% 
    group_by(Date,Creek) %>% 
    pivot_wider(names_from = Site, values_from = richness)
  
  # control creeks - we can bin together to make box plot
  p.controls <- p %>% 
    filter(Creek != "4Pad") %>% 
    mutate(rich_up_minus_down = Up - Dn) %>% 
    select(c(Date,Creek,rich_up_minus_down))
  
  # have to do padden separately to use up5 not up (ignore up11 for now)
  p.intervention <- p %>% 
    filter(Creek == "4Pad") %>% 
    mutate(rich_up_minus_down = Up5 - Dn) %>% 
    select(c(Date,Creek,rich_up_minus_down))
  
  pnew <- rbind(p.controls, p.intervention)
  
  return(pnew)
}

plot.delta.rich <- function(df){
  
  p <- df %>% 
  ggplot(aes(x=Date, y= rich_up_minus_down, fill=Creek, shape=Creek)) +
  scale_fill_manual(values = c("#669999", "#669999", "#669999","#CC0066", "#669999") ) +
  scale_shape_manual(values=c(21:25)) +
  geom_point(color="black", size=4) +
  geom_hline(yintercept = 0, linetype="dashed") +
  theme_bw() + 
  labs(y= "Richness Upstream - Richness Downstream", x= "Month")

return(p)
}

#ggsave(delta.rich.plot, filename = here("Output","figures","delta_asv_richness_atleast1biorep.png"))

delta.rich.123 <- calc.delta.rich(rich.123)
delta.rich.23 <- calc.delta.rich(rich.23)
delta.rich.3 <- calc.delta.rich(rich.3)

p1 <- plot.delta.rich(delta.rich.123) + theme(legend.position = "none") + ggtitle("COI, species level, found in any replicate")
p2 <- plot.delta.rich(delta.rich.23) + theme(legend.position = "none") + ggtitle("COI, species level, found in at least 2 replicates")
p3 <- plot.delta.rich(delta.rich.3) + ggtitle("COI, species level, found in all 3 replicates")

p1 + p2 + p3 + plot_layout(ncol = 3, nrow = 1, byrow = F)

```


## NMDS plots for sanity check    

Biological replicates really don't match up well. So do things even cluster by creek, month, and station? 

```{r nmds}

prepare.nmds <- function(df){
  p <- df %>%
    select(c(Sample_name,species_asv,reads)) %>% 
    pivot_wider(names_from = Sample_name, values_from = reads) %>% 
    column_to_rownames(var="species_asv")
  
  p[is.na(p)] <- 0
  
  p <- t(p)
  
  return(p)
}

visualize.nmds <- function(df,metadf){
  
  data.scores = as.data.frame(scores(df))
  meta <- metadf %>% 
    select(c(Sample_name, Date, Creek, Site)) %>% 
    distinct()
  data.scores$Sample = meta$Sample_name
  data.scores$Date = meta$Date
  data.scores$Creek = meta$Creek
  
  p <- ggplot(data.scores, aes(x = NMDS1, y = NMDS2)) + 
    geom_point(size = 4, aes(shape = Creek, fill= Date), colour = "black" )+ 
    labs(x = "NMDS1", fill = "Date", y = "NMDS2", shape = "Creek") +
    scale_shape_manual(values=21:25) +
    scale_fill_manual(values = c("#99FF00","#33CC99","#99FFFF","#3399FF","#CC99FF","#FF99CC")) +
    guides(fill=guide_legend(override.aes=list(colour=c("0321"="#99FF00","0421"="#33CC99", "0521"="#99FFFF", "0621"="#3399FF", "0721"="#CC99FF", "0821"="#FF99CC")))) +
    #scale_fill_manual(values = c("#99FF00","#33CC99","#3399FF","#FF99CC")) +
    #guides(fill=guide_legend(override.aes=list(colour=c("0321"="#99FF00","0421"="#33CC99",  "0621"="#3399FF",  "0821"="#FF99CC")))) +
    theme_bw() 
  
  return(p)
}

# # MIFISH remove barnes march bc something weird driving nmds
# data.123.mod <- data.123 %>% 
#   unite("creek_month", c("Creek", "Date"),sep="_", remove=FALSE) %>% 
#   filter(creek_month != "2Brn_0321") %>% 
#   select(-creek_month)
# data.23.mod <- data.23 %>% 
#   unite("creek_month", c("Creek", "Date"),sep="_", remove=FALSE) %>% 
#   filter(creek_month != "2Brn_0321") %>% 
#   select(-creek_month)
# data.3.mod <- data.3 %>% 
#   unite("creek_month", c("Creek", "Date"),sep="_", remove=FALSE) %>% 
#   filter(creek_month != "2Brn_0321") %>% 
#   select(-creek_month)



# COI: fuck at ASV level have to remove march and april bc they are RC'd

coi.123.nmds <- metaMDS(prepare.nmds(data.123), distance = "bray")
p.nmds.123 <- visualize.nmds(coi.123.nmds, data.123) + ggtitle("COI, ASV level, found in any replicate")
#ggsave(p.nmds.1, filename = here("Output","figures","COI_species_nmds_123biorep.png"))

coi.23.nmds <- metaMDS(prepare.nmds(data.23), distance = "bray")
p.nmds.23 <- visualize.nmds(coi.23.nmds, data.23) + ggtitle("COI, ASV level, found in at least 2 replicates")
#ggsave(p.nmds.1, filename = here("Output","figures","COI_species_nmds_23biorep.png"))

coi.3.nmds <- metaMDS(prepare.nmds(data.3), distance = "bray")
p.nmds.3 <- visualize.nmds(coi.3.nmds, data.3) + ggtitle("COI, ASV level, found in all 3 replicates")
#ggsave(p.nmds.1, filename = here("Output","figures","COI_species_nmds_3biorep.png"))

p.nmds.123 + p.nmds.23 + p.nmds.3 + plot_layout(ncol = 3, nrow = 1, byrow = F)

```

```{r nmds w species}

library(BiodiversityR) # also loads vegan
library(ggplot2)
library(readxl)
library(ggsci)
library(ggrepel)
library(ggforce)

BioR.theme <- theme(
        panel.background = element_blank(),
        panel.border = element_blank(),
        panel.grid = element_blank(),
        axis.line = element_line("gray25"),
        text = element_text(size = 12),
        axis.text = element_text(size = 10, colour = "gray25"),
        axis.title = element_text(size = 14, colour = "gray25"),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.key = element_blank())

meta <- data.123.mod %>% 
    select(c(Sample_name, Date, Creek, Site)) %>% 
    distinct()

Ordination.model1 <- metaMDS(prepare.nmds(data.123), distance='bray', k=2, trymax=1, 
  autotransform=TRUE, noshare=0.1, expand=TRUE, trace=1, plot=FALSE)
plot1 <- ordiplot(Ordination.model1, choices=c(1,2))

sites.long1 <- sites.long(plot1, env.data=meta)
head(sites.long1)

plotgg1 <- ggplot() + 
    geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
    geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
    xlab("NMDS1") +
    ylab("NMDS2") +  
    scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
    scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +    
    geom_point(data=sites.long1, 
               aes(x=axis1, y=axis2, colour=Creek, shape=Date), 
               size=5) +
    BioR.theme +
    ggsci::scale_colour_npg() +
    coord_fixed(ratio=1)

plotgg1

plotgg1 <- ggplot() + 
    geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
    geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
    xlab("NMDS1") +
    ylab("NMDS2") +  
    scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
    scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +    
    geom_point(data=sites.long(plot1, env.data=meta), 
               aes(x=axis1, y=axis2, colour=Creek, shape=Date), 
               size=5) +
    BioR.theme +
    ggsci::scale_colour_npg() +
    coord_fixed(ratio=1)

plotgg1


Ordination.model2 <- rda(prepare.nmds(data.123) ~ Creek, data=meta, scaling="species")
plot2 <- ordiplot(Ordination.model2, choices=c(1,2))
sites.long2 <- sites.long(plot2, env.data=meta)
head(sites.long1)
species.long2 <- species.long(plot2)
species.long2
axis.long2 <- axis.long(Ordination.model2, choices=c(1, 2))
axis.long2

plotgg2 <- ggplot() + 
    geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
    geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
    xlab(axis.long2[1, "label"]) +
    ylab(axis.long2[2, "label"]) +  
    scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
    scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +    
    geom_point(data=sites.long2, 
               aes(x=axis1, y=axis2, colour=Creek, shape=Date), 
               size=5) +
    geom_point(data=species.long2, 
               aes(x=axis1, y=axis2)) +
    BioR.theme +
    ggsci::scale_colour_npg() +
    coord_fixed(ratio=1)
plotgg2

spec.envfit <- envfit(plot2, env=prepare.nmds(data.123))
spec.data.envfit <- data.frame(r=spec.envfit$vectors$r, p=spec.envfit$vectors$pvals)
species.long2 <- species.long(plot2, spec.data=spec.data.envfit)
species.long2

species.long3 <- species.long2[species.long2$p <= 0.001, ]
species.long3

plotgg2 <- ggplot() + 
    geom_vline(xintercept = c(0), color = "grey70", linetype = 2) +
    geom_hline(yintercept = c(0), color = "grey70", linetype = 2) +  
    xlab(axis.long2[1, "label"]) +
    ylab(axis.long2[2, "label"]) +  
    scale_x_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +
    scale_y_continuous(sec.axis = dup_axis(labels=NULL, name=NULL)) +    
    geom_point(data=sites.long2, 
               aes(x=axis1, y=axis2, colour=Creek, shape=Date), 
               size=5) +
    geom_segment(data=species.long3, 
                 aes(x=0, y=0, xend=axis1*4, yend=axis2*4), 
                 colour="red", size=0.7, arrow=arrow()) +
    geom_text_repel(data=species.long3, 
                    aes(x=axis1*4, y=axis2*4, label=labels),
                    colour="red") +
    BioR.theme +
    ggsci::scale_colour_npg() +
    coord_fixed(ratio=1)

plotgg2

```